Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-04 ind_files/04_field_eqns.tex
Minor rephrasing

Equation \eqref{cts_replicator_mutator} also recovers~\posscite{kimura_stochastic_1965} continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming\footnote{This actually need not be an assumption, and can be derived as a theorem under biologically reasonable conditions on the birth/death functionals upon taking the rigorous approach to both the system size expansion (rescaling) and the infinite population limit. See, for example,~\citet{etheridge_introduction_2000} or proposition 4.1 in~\citet{champagnat_individual_2008}} that even though the total number of individuals $\int \nu(x) dx \to \infty$ as $K \to \infty$, the \emph{scaled} population size $N_K = \int \nu(x) dx / K$ remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have
\begin{equation*}
\int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}}|m(y-z)\psi(z,t)|dydz =  \int\limits_{\mathbb{R}}\left(\int\limits_{\mathbb{R}}|m(y-z)|dy\right)|\psi(z,t)|dz
= \|m\|_{L^{1}(\mathbb{R})}\|\psi\|_{L^{1}(\mathbb{R})} < \infty
\end{equation*}
We can therefore conclude that $m(y-z)\psi(z,t) \in  L^{1}(\mathbb{R}\times\mathbb{R})$ for any given $t>0$, meaning we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

>>>>

Equation \eqref{cts_replicator_mutator} also recovers~\posscite{kimura_stochastic_1965} continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming\footnote{This actually need not be an assumption, and can be derived as a theorem under biologically reasonable conditions on the birth/death functionals upon taking the rigorous approach to both the system size expansion (rescaling) and the infinite population limit. See, for example,~\citet{etheridge_introduction_2000} or proposition 4.1 in~\citet{champagnat_individual_2008}} that even though the total number of individuals $\int \nu(x) dx \to \infty$ as $K \to \infty$, the \emph{scaled} population size $N_K = \int \nu(x) dx / K$ remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have
\begin{align*}
\int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}}|m(y-z)\psi(z,t)|dydz &=  \int\limits_{\mathbb{R}}\left(\int\limits_{\mathbb{R}}|m(y-z)|dy\right)|\psi(z,t)|dz\\
&=\left(\int\limits_{\mathbb{R}}|m(x)|dx\right)\left(\int\limits_{\mathbb{R}}|\psi(z,t)|dz\right) < \infty
\end{align*}
We can therefore conclude that $m(y-z)\psi(z,t) \in  L^{1}(\mathbb{R}\times\mathbb{R})$ for any given $t>0$, meaning we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

### 2023-04-04 ind_files/05_unification_discussion.tex
Minor rephrasing

The $\overline{\tau}\sigma^2_{f}$ term quantifies the loss of variation due to stochastic extinctions (i.e. demographic stochasticity) and thus represents the classic effect of neutral genetic (or ecological) drift in finite populations. To see this, it is instructive to consider the case in which this is the only force at play. Let us imagine a population of asexual organisms in which each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observational study and subsequently passed on to offspring - For example, a neutral genetic tag in a part of the genome that experiences a negligible mutation rate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\partial f/\partial t,f\right) = 0$. From \eqref{nD_stochastic_Price_variance}, we see that in this case, the variance changes as
\begin{equation}
d\sigma^2_f = - \frac{\overline{\tau}\sigma^2_{f}}{KN_K}dt + \frac{1}{\sqrt{K}N_{K}(t)}dW_{\sigma^2_{f}}
\end{equation}
On taking expectations, the second term on the RHS vanishes and we see that the expected variance in the population obeys
\begin{equation}
\label{neutral_example_for_variances}
\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]
\end{equation}
where I have decomposed the expectation of the product on the RHS into a product of expectations, which is admissible since the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. Equation \eqref{neutral_example_for_variances} is a simple linear ODE and has the solution
\begin{equation}
\mathbb{E}[\sigma^2_f](t) = \sigma^2_f(0)e^{-\mathbb{E}\left[\frac{\overline{\tau}}{KN_K}\right]t}
\end{equation}
which tells us that the expected diversity (variance) of labels in the population decreases exponentially over time. The rate of loss is $\mathbb{E}\left[\overline{\tau}(KN_K)^{-1}\right]$, and thus, populations with higher mean turnover $\overline{\tau}$  and/or lower population size $KN_K$ lose diversity faster. This is because populations with higher $\overline{\tau}$ experience more stochastic events per unit time (a gambler's ruin type scenario), while extinction is `easier' in smaller populations because a smaller number of deaths is required to eliminate a label from the population completely. Note that \emph{which} labels/individuals are lost is entirely random (since all labels are arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.

>>>>

The $\overline{\tau}\sigma^2_{f}$ term quantifies the loss of variation due to stochastic extinctions (i.e. demographic stochasticity) and thus represents the classic effect of neutral genetic (or ecological) drift in finite populations. To see this, it is instructive to consider the case in which this is the only force at play. Let us imagine a population of asexual organisms in which each $f_i$ is simply a label or mark arbitrarily assigned to individuals in the population at the start of an experiment/observational study and subsequently passed on to offspring - For example, a neutral genetic tag in a part of the genome that experiences a negligible mutation rate. Let us set $\mu = 0$ so that the labels cannot change between parents to offspring. This means that we have $M_{\sigma^2_f}(\mathbf{p},N_K) \equiv 0$. Further, since the labels are arbitrary and have no effect whatsoever on the biology of the organisms, each label has the same fitness $w_i \equiv w$ and per-capita turnover $\tau_i \equiv \tau$, and thus $\overline{w} = w$ and $\overline{\tau} = \tau$. Note that since every type has the same fitness and turnover rate, we have $\textrm{Cov}\left(w,(f - \overline{f})^2\right) \equiv \textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \equiv 0$. Since the labels do not change over time, we also have $\textrm{Cov}\left(\partial f/\partial t,f\right) = 0$. From \eqref{nD_stochastic_Price_variance}, we see that in this case, the variance changes as
\begin{equation}
d\sigma^2_f = - \frac{\tau\sigma^2_{f}}{KN_K}dt + \frac{1}{\sqrt{K}N_{K}(t)}dW_{\sigma^2_{f}}
\end{equation}
Where I have written $\overline{\tau} = \tau$, since every type has the same turnover rate. On taking expectations, the second term on the RHS vanishes and we see that the expected variance in the population obeys
\begin{equation}
\label{neutral_example_for_variances}
\frac{d \mathbb{E}[\sigma^2_f]}{dt} = - \left(\mathbb{E}\left[\frac{\tau}{KN_K}\right]\right)\mathbb{E}[\sigma^2_{f}]
\end{equation}
where I have decomposed the expectation of the product on the RHS into a product of expectations, which is admissible since the label $f$ is completely arbitrary and thus independent of both $\overline{\tau}$ and $N_K(t)$. Equation \eqref{neutral_example_for_variances} is a simple linear ODE and has the solution
\begin{equation}
\mathbb{E}[\sigma^2_f](t) = \sigma^2_f(0)e^{-\mathbb{E}\left[\frac{\tau}{KN_K}\right]t}
\end{equation}
which tells us that the expected diversity (variance) of labels in the population decreases exponentially over time. The rate of loss is $\mathbb{E}\left[\tau(KN_K)^{-1}\right]$, and thus, populations with higher turnover rate $\tau$  and/or lower population size $KN_K$ lose diversity faster. This is because populations with higher $\tau$ experience more stochastic events per unit time (a gambler's ruin type scenario), while extinction is `easier' in smaller populations because a smaller number of deaths is required to eliminate a label from the population completely. Note that \emph{which} labels/individuals are lost is entirely random (since all labels are arbitrary), but nevertheless, labels tend to be stochastically lost until only a small number of labels remain in the population.
